---
title: "Tutorial4 : Variable selection "
author: "Tanakamol Mahawan"
date: "2023-12-12"
output: html_document
---

# Introduction

Variable selection is the process of selecting the most relevant features for a machine learning model. It’s crucial because it can enhance the model’s prediction performance, provide faster and more cost-effective predictions, and offer a better understanding of the data. 

We use combination of methods : LASSO selection first (100 models) , then selected genes from LASSO will be further run in Boruta and VarselRF(100 models), The overlapping genes of these 2 methods will be finally selected for modelling. 

Selection criteria : Genes must be found in at least 80% and 5 splits out of 10,

# 1.Loading Required Libraries
Before we begin, make sure to install and load the required packages:

```{r load_libraries, message=FALSE, warning=FALSE}
#BiocManager::install("MultiBaC")

library(smotefamily)
library(MLmetrics)
library(mlr3measures)
library(clusterProfiler)
library(org.Hs.eg.db)
#Load all custom functions
source("custom_functions.r")
```

# 2.Loading data and phenotype

```{r ,echo=FALSE}
# Loads all data using in gene filtering
load("Data_after_integration.rdata")
```

```{r ,echo=FALSE}
abc<- ABC_data
abcMeta<-ABC_pheno
```

```{r , eval=FALSE}
abc <- read.csv("arsyn1ABC_rmOLE.csv",row.names = 1, header = T)
abcMeta <- read.csv("pheno_Arsyn1ABC.csv",row.names = 1, header = T)

```

```{r }
#Check sample and class distribution 
table(abcMeta$SAMPLE)
table(abcMeta$CLASS)

#prep variables for context
#find the match for the ensembl ids
tableIDs<-bitr(colnames(abc),fromType = "ENSEMBL",toType = "SYMBOL",OrgDb = org.Hs.eg.db)
head(tableIDs)
```

# 3.Check data by PCA plot 

```{r }
#2. check data ----

PCA_plot(abc,abcMeta$CLASS, title = "ABC data")

```

# 4.Preparing data splits

Split data into 10 splits 
```{r , results='hide'}
# Split train 90% and test 10% 
abcSplits<-applyPrepData(data_train = abc,prepData = prepData,factor_var = factor(abcMeta$CLASS),split = 0.9,n = 10)
```
# 5.Apply the variable selection algorithms to each of the splits

LASSO is used first.
```{r, eval=FALSE}

abc_LassoRes<-applyMultiVarSelectionForCrossVal_withModel(inputListwSplits = abcSplits,functionSelection = TwoClassLasso,thresMod = 80,ntree = 500 )


abc_LassoRes_concRes<-as.data.frame(table(concatenateVarSelec(abc_LassoRes)))
abc_LassoRes_concRes$GeneSymbol<-tableIDs[match(abc_LassoRes_concRes$Var1,tableIDs[,1]),2]

#View(abc_LassoRes_concRes)

write.csv(abc_LassoRes_concRes,"abcLassoResults_FrequencyTable_relax.csv",row.names = F)

```

```{r}
head(abc_LassoRes_concRes)
```

# 6.Modelling and extracting model metrics of LASSO 

```{r , eval= FALSE}
abc_LassoRes_modelMetExtract<-modelMetrExtract(abc_LassoRes)

View(abc_LassoRes_modelMetExtract)

write.csv(abc_LassoRes_modelMetExtract,"abcLassoResultsModel_10DataSplits_relax.csv")

```

# 7.Plot of LASSO selected genes
```{r}
#lasso
selectedGenes <- abc_LassoRes_concRes[abc_LassoRes_concRes$Freq>5,1]
length(selectedGenes) #38 genes

PCA_plot(abc[,colnames(abc)%in%selectedGenes],abcMeta$CLASS, title = "Lasso selected genes (n=38)")

bp_lasso <- boxplotSubsetPDACmetastasis(soma = abc,selVars = selectedGenes[1:9],metadata = abcMeta,tableIDs = tableIDs)
#boxplots showing first 9 genes 
```

#  Backward selection after LASSO selection  

# 8.Apply the backwards selection algorithms to each of the splits 

We use selected genes from LASSO to further split and select by backwards selection called varselRF. 

```{r, eval=FALSE}

abcSplits_afterLasso<-applyPrepData(data_train = abc[,colnames(abc)%in%selectedGenes],prepData = prepData,factor_var = factor(abcMeta$CLASS),split = 0.9,n = 10)

abc_BackwardsAfterLasso <- applyMultiVarSelectionForCrossVal_withModel(
  inputListwSplits = abcSplits_afterLasso,
  functionSelection = BackSelec_wTimes,
  thresMod = 80,
  ntree = 500
)

abc_Backwards_concRes<-as.data.frame(table(concatenateVarSelec(abc_BackwardsAfterLasso)))
abc_Backwards_concRes$GeneSymbol<-tableIDs[match(abc_Backwards_concRes$Var1,tableIDs[,1]),2]
View(abc_Backwards_concRes)
write.csv(abc_Backwards_concRes,"abc_Backwards_FrequencyTable.csv",row.names = F)

```

```{r}
selectedGenes_Backwards<-abc_Backwards_concRes[abc_Backwards_concRes$Freq>=5,1]
length(selectedGenes_Backwards) #15 genes

selectedGenes_df <- data.frame(selectedGenes_Backwards)
selectedGenes_df$GeneSymbol <-tableIDs[match(selectedGenes_Backwards,tableIDs[,1]),2]
print(selectedGenes_df)
```

# 9.Modelling of Backwards selected genes 
```{r ,eval=FALSE}

abc_BackwardsRes_modelMetExtract<-modelMetrExtract(abc_BackwardsAfterLasso)
View(abc_BackwardsRes_modelMetExtract)
write.csv(abc_BackwardsRes_modelMetExtract,"abc_BackwardsResmodel_10DataSplits.csv")
```

# 10.Plot of Backwards selected genes

```{r }

PCA_plot(abc[,colnames(abc)%in%selectedGenes_Backwards],abcMeta$CLASS,
         title = "Backwards selected genes(n=15)")

bp_Backwards <- boxplotSubsetPDACmetastasis(soma = abc,selVars = selectedGenes_Backwards[1:9],metadata = abcMeta,tableIDs = tableIDs)
#boxplots showing first 9 genes 
```         

# Boruta after LASSO selection

# 11.apply the Boruta selection algorithms to each of the splits 

```{r ,eval=FALSE}
abc_Boruta_afterLasso <- applyMultiVarSelectionForCrossVal_withModel(
  inputListwSplits = abcSplits_afterLasso,
  functionSelection = Boruta_wTimes,
  thresMod = 80,
  ntree = 500
)


abc_Boruta_concRes<-as.data.frame(table(concatenateVarSelec(abc_Boruta_afterLasso)))

abc_Boruta_concRes$GeneSymbol<-tableIDs[match(abc_Boruta_concRes$Var1,tableIDs[,1]),2]

View(abc_Boruta_concRes)

write.csv(abc_Boruta_concRes,"abc_Boruta_FrequencyTable.csv",row.names = F)

```


```{r}
head(abc_Boruta_concRes)
```
# 12.Modelling of Boruta  selected genes 

```{r , eval=FALSE}
#Boruta
abc_BorutaRes_modelMetExtract<-modelMetrExtract(abc_Boruta_afterLasso )
View(abc_BorutaRes_modelMetExtract)
write.csv(abc_BorutaRes_modelMetExtract,"abc_BorutaResmodel_10DataSplits.csv")
```

# 13.Some plots of boruta selected genes

```{r}

selectedGenes_Boruta <- abc_Boruta_concRes[abc_Boruta_concRes$Freq>5,1]
#22 genes 

PCA_plot(abc[,colnames(abc)%in%selectedGenes_Boruta],abcMeta$CLASS,title = "Boruta selected genes(n=22)")

bp_Boruta<- boxplotSubsetPDACmetastasis(soma = abc,selVars = selectedGenes_Boruta[1:9],metadata = abcMeta,tableIDs = tableIDs)
#boxplots showing first 9 genes 
```

# 14.Overlap of selectedGenes_Boruta and selectedGenes_Backwards

The overlapping genes will be used in modelling.
```{r}

Overlap_genes <- intersect(selectedGenes_Backwards,selectedGenes_Boruta )
length(Overlap_genes) # 15 genes (all backwards genes are found in boruta)
Overlap_genes <- abc_Backwards_concRes[abc_Backwards_concRes$Var1%in%Overlap_genes,]
Overlap_genes <- Overlap_genes[order(Overlap_genes$Freq, decreasing = T),]
print(Overlap_genes)

```

#Save results
```{r, eval=FALSE}
write.csv(Overlap_genes,"Overlap_genes.csv")
save.image("4.variable_selection.rdata")
```

