---
title: "Tutorial2 : Gene filtering and analysing gene expression data "
author: "Tanakamol Mahawan"
date: "2023-12-12"
output: html_document
---

# Introduction

In this tutorial, we will guide you through the process of filtering and analysing gene expression data from multiple datasets (TCGA, AUS, and CA). The objective is to identify genes with consistent fold change direction and eliminate noise.

# 1.Loading Required Libraries

```{r load_libraries, message=FALSE, warning=FALSE}
library(stats)
library(dplyr)
library(plyr)
library(Biobase)
#Load all custom functions
source("custom_functions.r")
```


# 2.Load data 

```{r , echo=FALSE}
# Loads all data using in gene filtering
load("Data_before_integration.rdata")
```

```{r, eval = FALSE}

# Read  data and phenotype information
pheno<- read.csv("phenotype_all_5data.csv",header = T, stringsAsFactors = F)
head(pheno)

TCGA <- read.csv('./normLog_TCGA.csv',header=TRUE,stringsAsFactors = F,row.names = 1) 
TCGA$class <- as.factor(TCGA$class)
pheno_TCGA <- pheno[pheno$SAMPLE=="A",c(1,3,5,6,7)]
head(pheno_TCGA)

AUS <- read.csv('./normLog_AUS.csv',header=TRUE,stringsAsFactors = F,row.names = 1) 
AUS$class <- as.factor(AUS$class)
pheno_AUS <- pheno[pheno$SAMPLE=="B",c(1,3,5,6,7)]
head(pheno_AUS)

CA <- read.csv('./normLog_CA.csv',header=TRUE,stringsAsFactors = F,row.names = 1) #read
CA$class <- as.factor(CA$class)
pheno_CA <- pheno[pheno$SAMPLE=="C",c(1,3,5,6,7)]
head(pheno_CA)

# Transpose and convert data frames
TCGA <- as.data.frame(t(TCGA[, -1]))
AUS <- as.data.frame(t(AUS[, -1]))
CA <- as.data.frame(t(CA[, -1]))

```

# 3.Find Overlapping Gene IDs

Identifying the common set of genes across datasets is crucial for comparative analysis and downstream processing.

```{r }

# Find the overlap of gene IDs across the three datasets
ABC_names <- Reduce(intersect, list(row.names(TCGA),
                                    row.names(AUS),
                                    row.names(CA)))
paste0("#overlapping genes :",length(ABC_names)) #How many overlapping genes?

# Filter Data Based on Overlapping Gene IDs

TCGA <- TCGA[ row.names(TCGA) %in% ABC_names  , ]
dim(TCGA)

AUS <- AUS[ row.names(AUS) %in% ABC_names  , ]
dim(AUS)

CA <-  CA[ row.names(CA) %in% ABC_names  , ]
dim(CA)
```

# 4. Calculate Mean Expression

```{r }
# Calculate the mean expression for each gene across different classes in each dataset
# TCGA
TCGA_mean <- aggregate(t(TCGA),
                by = list(pheno_TCGA$CLASS),
                FUN = mean)

TCGA_mean <-  setNames(data.frame(t(TCGA_mean[,-1])), TCGA_mean[,1])
TCGA_mean$FC <- TCGA_mean$metastasis- TCGA_mean$non_metastasis

#AUS
AUS_mean <- aggregate(t(AUS),
                       by = list(pheno_AUS$CLASS),
                       FUN = mean)
AUS_mean <-  setNames(data.frame(t(AUS_mean[,-1])), AUS_mean[,1])
AUS_mean$FC <- AUS_mean$metastasis- AUS_mean$non_metastasis

#CA
CA_mean <- aggregate(t(CA),
                       by = list(pheno_CA$CLASS),
                       FUN = mean)
CA_mean <-  setNames(data.frame(t(CA_mean[,-1])), CA_mean[,1])
CA_mean$FC <- CA_mean$metastasis- CA_mean$non_metastasis

# Combine fold changes from different datasets into a single dataframe

FC_all <- as.data.frame(cbind(rownames(TCGA_mean),TCGA_mean$FC,AUS_mean$FC,CA_mean$FC ))
names(FC_all)[] <- c("gene","TCGA","AUS","CA")
head(FC_all)

```

# 5. Direction of Fold Change

Considering the direction of fold change helps filter genes that consistently change in the same direction across datasets.

```{r }
# Convert relevant columns to numeric
i <- c(2:4)
FC_all[ , i] <- apply(FC_all[ , i], 2, function(x) as.numeric(as.character(x)))
sapply(FC_all, class)  

# Calculate the direction of fold change for each dataset
FC_all$TCGA2 <- FC_all$TCGA / abs(FC_all$TCGA)
FC_all$AUS2 <- FC_all$AUS / abs(FC_all$AUS)
FC_all$CA2 <- FC_all$CA / abs(FC_all$CA)
FC_all$sum <- FC_all$TCGA2+FC_all$AUS2+FC_all$CA2
FC_all$abs <- abs(FC_all$sum)
head(FC_all)
```

# 6. Filter Only Genes with the Same Direction

```{r }
# Filter genes that have the same direction of fold change in all three datasets
fil_genes <- FC_all[FC_all$abs == 3, ]
head(fil_genes)
```

# 7. Save the results 

fil_genes is gene lists that have save expression patterns ,  we will use this in next step which is data integration and variable selection. 

```{r }
# Save the final filtered fold change data
write.csv(fil_genes, "FC_filtered_ABC.csv")
# Save the R environment
save.image("2_ilter_genes.rdata")

```

